`{traveltime}`: an R package to calculate walking and motorised travel time from user-specified locations

### 2025-01-29 paper/paper.qmd
paper: polishing / tidying of ms and repo

---
title: "`{traveltime}`: an R package to calculate walking and motorised travel time from user-specified locations"
format:
  #md: default
  html:
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nick Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 3 "
  - name: Daniel Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
date: In prep 2025
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and a myriad others [@nelson2019suite]. Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref]. `{traveltime}` enables a user to supply a set of geographic point locations and an extent of interest, to then calculate the motorised or walking travel time over that extent. The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.

@@

`{traveltime}` provides a spatial interface using object classes from the `terra` package `[@terra]`. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and extents as `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` objects; and returns the result as a `SpatRaster` object. The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest. The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global available via the `malariaAtlas` package [@pfeffer2018malariaatlas]. Rather than house and manipulate these surfaces independently, `{traveltime}` relies on `malariaAtlas` to download the surfaces for the extent of interest.

@@

Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use (collectively >1500 citations per google scholar at the 28th of January 2025), and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata]. There is clear demand for these type of products.

@@

Other R packages produce superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the <https://www.TravelTime.com> API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repetitive iteration for all combinations of time and points, and additional calculations of the minimum value for each cell from all points. Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use.

@@

-   our extent of interest --- in this case a map of Singapore, and
-   our points to calculate travel time from --- here the locations of Singapore's MRT and LRT stations.

@@

The the `stations` data set included in the `{traveltime}` package is a 563 \$\*\$ 2 `matrix` containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata.

@@

Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest. We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.)

We're also only interested in walking *on land* so we mask out areas outside of `sin`, that are within the extent of the raster:

@@

We present the resulting calculated travel times in Figure \ref{fig:data}

```{r plot result, fig.cap = "\\label{fig:result}Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."}
ggplot() +
  geom_spatraster(
    data = travel_time_sin
  ) +
  scale_fill_viridis_c(
    option = "A",
    direction = -1,
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "minutes") +
  geom_spatvector(
    data = sin,
    fill = "transparent",
    col = "grey20"
  )
```

Note that the above raster `travel_time_sin` includes infinite (`Inf`) values. In Figure \ref{fig:data}, a number of islands to the south are shown as cells unconnected with the mainland. These raster cells for these islands appear absent in Figure \ref{fig:result}. Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig:result}.

@@

Firstly, `{traveltime}` currently has access to friction surfaces for 2020, both at 30 arc-second resolution. As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a 5 km cell size is a relatively large area to walk across, and thus actual waking times are likley to vary significantly within each cell.

At the other end of the scale, the package can require relatively large amounts of onboard memory for analyses over large landscapes (e.g. one analyses over Africa required \~ 72 GB RAM). Developing methods to handle large landscapes either with less memory or via cloud resources would be helpful to make such analyses accessible to those without access to larger computing resources.

>>>>

---
title: "`{traveltime}`: an R package to calculate walking and motorised travel time from user-specified locations"
format:
  #md: default
  html:
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E. Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 3 "
  - name: Daniel J. Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
date: In prep 2025
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref]. `{traveltime}` enables a user to calculate the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates. The result is a raster of the area of interest, of 30 arcsecond resolution[^1] where the value in each cell is the lowest travel time in minutes to any of the specified locations.

[^1]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator

@@

`{traveltime}` provides a spatial interface using object classes from the `terra` package `[@terra]`. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object. The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest. The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global and available from the `malariaAtlas` package [@pfeffer2018malariaatlas]. Rather than house and manipulate these surfaces independently, `{traveltime}` relies internally on `malariaAtlas` to download the surfaces for the area of interest.

@@

Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use[^2], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata]. There is clear demand for these type of products.

[^2]: Collectively >1600 citations per Google Scholar at the 28th of January 2025.

@@

Other R packages produce superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, and additional calculations of the minimum value for each cell from all points. Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.

@@

-   our area of interest --- in this case a map of Singapore, and
-   our points to calculate travel time from --- here the locations of Singapore's MRT and LRT stations.

@@

The the `stations` data set included in the `{traveltime}` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from @singdata.

@@

Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest. We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we mask out areas outside of `sin`, that are within the extent of the raster:

@@

We present the resulting calculated travel times in Figure \ref{fig:result}. Note that the results in `travel_time_sin` include infinite (`Inf`) values. In Figure \ref{fig:data}, the islands to the south and north-east are shown as filled cells, but unconnected with the mainland. These raster cells for these islands appear absent in Figure \ref{fig:result}. Because they are not connected to any cells with a station, the calculated travel time is infinite, and so these cells do not appear in Figure \ref{fig:result}.

```{r plot result, fig.cap = "\\label{fig:result}Map of walking travel time in Singapore, in minutes from nearest MRT or LRT station."}
ggplot() +
  geom_spatraster(
    data = travel_time_sin
  ) +
  scale_fill_viridis_c(
    option = "A",
    direction = -1,
    na.value = "transparent"
  ) +
  theme_minimal() +
  labs(fill = "Minutes") +
  geom_spatvector(
    data = sin,
    fill = "transparent",
    col = "grey20"
  )
```

@@

Firstly, `{traveltime}` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution. As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km$^2$ cell size is a relatively large area to walk across, and thus actual waking times are likley to vary significantly within each cell. We note though that the capability to calculate travel time over any friction surface exists presently, it is only the availability of additional surfaces that is presently restrictive.

At the other end of the scale, the calculations can require relatively large amounts of onboard memory for analyses over large landscapes (e.g. one analyses over Africa required \~ 72 GB RAM). Developing methods to handle large landscapes either with less memory or via cloud resources would be helpful to make such analyses accessible to those without access to larger computing resources.
